---
title: "Model Forms: Mathematics of the Five curveRcore Calibration Models"
subtitle: "Forward functions, inverses, derivatives, gradient factory, and prediction grids"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 7
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Model Forms: Mathematics of the Five curveRcore Calibration Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: core_references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center"
)

# From GitHub:
# devtools::install_github("immunoplex/curveRcore")

library(curveRcore)
```

---

# Introduction

This vignette is a self-contained mathematical reference for the five
sigmoidal calibration-curve models implemented in **curveRcore**.
It documents the forward functions, their analytical inverses, first and
second derivatives, the gradient-factory used for delta-method error
propagation, and the prediction-grid utilities shared by
**curveRfreq** and **curveRbayes**.

No model fitting is performed here.
The audience is a statistician or assay scientist who wants to understand
the mathematics before—or independently of—running a full calibration
workflow.
For a worked end-to-end analysis, see the companion vignette
*Calibration Curve Workflows*.

## Notation and conventions

All five models share the following parameter convention, enforced
across the entire curveR ecosystem:

| Symbol | Role | Constraint |
|--------|------|-----------|
| $a$ | Lower asymptote (baseline response) | $a < d$ |
| $b$ | Scale / slope parameter | $b > 0$ |
| $c$ | Inflection-point location on the $x$-axis | real |
| $d$ | Upper asymptote (saturation response) | $d > a$ |
| $g$ | Asymmetry parameter (5-parameter models only) | $g > 0$ |
| $x$ | Independent variable passed to the model | model-dependent |

The independent variable $x$ is **scale-agnostic**: any monotone
transformation (e.g.\ $x = \log_{10}(\text{concentration})$) is applied
*upstream* in R before the model is called.
The models themselves do not perform any transformation internally,
except `loglogistic4` and `loglogistic5`, which require $x > 0$ on the
raw (un-logged) concentration scale.

---

# Forward Model Equations {#sec-forward}

## Four-Parameter Logistic (4PL) — `logistic4`

$$
y = a + \frac{d - a}{1 + \exp\!\left(-\dfrac{x - c}{b}\right)}
\tag{4PL}
$$

The 4PL is symmetric about its inflection point $(c,\,(a+d)/2)$.
It is the workhorse model for log-linear immunoassay standard curves.

```{r forward-4pl}
# Parameter set used throughout this vignette
p4 <- list(a = 100, b = 0.8, c = 1.5, d = 50000)

x_seq <- seq(-1, 4, length.out = 300)
y_4pl <- logistic4(x_seq, p4$a, p4$b, p4$c, p4$d)

plot(x_seq, y_4pl, type = "l", lwd = 2, col = "#2166AC",
     xlab = expression(log[10](concentration)),
     ylab = "Response (MFI)",
     main = "4PL forward curve")
abline(v = p4$c, h = (p4$a + p4$d) / 2, lty = 2, col = "grey50")
legend("topleft", legend = "logistic4", lwd = 2, col = "#2166AC", bty = "n")
```

## Five-Parameter Logistic (5PL) — `logistic5`

$$
y = a + \frac{d - a}{\left(1 + \exp\!\left(-\dfrac{x - c}{b}\right)\right)^{g}}
\tag{5PL}
$$

When $g = 1$ this reduces exactly to the 4PL.
$g > 1$ compresses the lower shoulder (skews toward the upper asymptote);
$0 < g < 1$ compresses the upper shoulder.
The inflection point shifts with $g$:

$$
x_{\inf} = c + b\ln(g), \qquad
y_{\inf} = a + (d-a)\left(\frac{g}{g+1}\right)^{g}
$$

```{r forward-5pl}
p5 <- list(a = 100, b = 0.8, c = 1.5, d = 50000, g = 0.6)

y_5pl <- logistic5(x_seq, p5$a, p5$b, p5$c, p5$d, p5$g)

plot(x_seq, y_4pl, type = "l", lwd = 2, col = "#2166AC", lty = 2,
     xlab = expression(log[10](concentration)),
     ylab = "Response (MFI)",
     main = "4PL vs 5PL (g = 0.6)")
lines(x_seq, y_5pl, lwd = 2, col = "#D6604D")
legend("topleft",
       legend = c("logistic4 (g = 1)", "logistic5 (g = 0.6)"),
       lwd = 2, col = c("#2166AC", "#D6604D"), lty = c(2, 1), bty = "n")
```

## Four-Parameter Log-Logistic — `loglogistic4`

$$
y = a + \frac{d - a}{1 + (c / x)^{b}}
\tag{LL4}
$$

Here $c > 0$ is the EC50 on the *raw* concentration scale, and $b$ is the
Hill coefficient.
This is the classical Hill equation / dose-response parameterisation.
**Domain restriction:** requires $x > 0$.
When working on the $\log_{10}$ scale, `logistic4` is algebraically
equivalent and preferred.

```{r forward-ll4}
p_ll4 <- list(a = 100, b = 1.8, c = 30, d = 50000)  # c is EC50 in AU/mL
x_conc <- seq(0.01, 500, length.out = 500)

y_ll4 <- loglogistic4(x_conc, p_ll4$a, p_ll4$b, p_ll4$c, p_ll4$d)

plot(log10(x_conc), y_ll4, type = "l", lwd = 2, col = "#4DAC26",
     xlab = expression(log[10](concentration)),
     ylab = "Response (MFI)",
     main = "Log-logistic 4PL — concentration scale")
abline(v = log10(p_ll4$c), lty = 2, col = "grey50")
```

## Five-Parameter Generalised Logistic (Richards) — `loglogistic5`

$$
y = a + (d - a)\,\bigl(1 + g\,\exp(-b(x - c))\bigr)^{-1/g}
\tag{LL5}
$$

This is the Richards / generalised logistic curve [@Richards1959].
Despite the name `loglogistic5`, $x$ here is the *log-transformed*
concentration (i.e.\ the model operates on the log scale).
When $g \to 0$ the formula approaches the Gompertz; when $g = 1$ it
matches the 4PL up to a parameter rescaling.

```{r forward-ll5}
p_ll5 <- list(a = 100, b = 1.2, c = 1.5, d = 50000, g = 0.5)

y_ll5 <- loglogistic5(x_seq, p_ll5$a, p_ll5$b, p_ll5$c, p_ll5$d, p_ll5$g)

plot(x_seq, y_ll5, type = "l", lwd = 2, col = "#762A83",
     xlab = expression(log[10](concentration)),
     ylab = "Response",
     main = "loglogistic5 (Richards / generalised logistic)")
```

## Four-Parameter Gompertz — `gompertz4`

$$
y = a + (d - a)\,\exp\!\bigl(-\exp(-b(x - c))\bigr)
\tag{G4}
$$

The Gompertz is intrinsically asymmetric: it rises steeply from the lower
asymptote and approaches the upper asymptote more gradually
[@Gompertz1825].
The inflection occurs at $x = c$, where $y_{\inf} = a + (d-a)/e$.

```{r forward-gompertz}
p_g4 <- list(a = 100, b = 1.2, c = 1.5, d = 50000)

y_g4 <- gompertz4(x_seq, p_g4$a, p_g4$b, p_g4$c, p_g4$d)

plot(x_seq, y_4pl,  type = "l", lwd = 2, col = "#2166AC", lty = 2,
     xlab = expression(log[10](concentration)),
     ylab = "Response",
     main = "4PL vs Gompertz — asymmetry comparison")
lines(x_seq, y_g4, lwd = 2, col = "#BF812D")
abline(v = p_g4$c, lty = 3, col = "grey60")
legend("topleft",
       legend = c("logistic4 (symmetric)", "gompertz4 (asymmetric)"),
       lwd = 2, col = c("#2166AC", "#BF812D"), lty = c(2, 1), bty = "n")
```

## Summary comparison table

| Model | Equation kernel | Symmetry | Parameters | `x` domain |
|-------|----------------|----------|-----------|------------|
| `logistic4` | $(1 + e^{-(x-c)/b})^{-1}$ | Symmetric | $a,b,c,d$ | $\mathbb{R}$ |
| `logistic5` | $(1 + e^{-(x-c)/b})^{-g}$ | Asymmetric | $a,b,c,d,g$ | $\mathbb{R}$ |
| `loglogistic4` | $(1 + (c/x)^b)^{-1}$ | Symmetric | $a,b,c,d$ | $x > 0$ |
| `loglogistic5` | $(1 + g\,e^{-b(x-c)})^{-1/g}$ | Asymmetric | $a,b,c,d,g$ | $\mathbb{R}$ |
| `gompertz4` | $\exp(-e^{-b(x-c)})$ | Asymmetric | $a,b,c,d$ | $\mathbb{R}$ |

---

# Analytical Inverse Functions {#sec-inverses}

Each model has a closed-form inverse that maps an observed response $y$
back to the concentration scale.
Two variants exist for every model:

- **`inv_<model>(y, a, b, c, d)`** — $a$ is a free (estimated) parameter.
- **`inv_<model>_fixed(y, fixed_a, b, c, d)`** — $a$ is supplied as an
  external constant (e.g.\ from a plate blank mean), reducing Monte Carlo
  uncertainty.

All functions return `NA` for $y$ values outside the open interval
$(a, d)$.

## Inverse formulae

### 4PL inverse

Solving $y = a + (d-a)/(1 + e^{-(x-c)/b})$ for $x$:

$$
x = c + b\,\ln\!\frac{y - a}{d - y}
\tag{inv-4PL}
$$

### 5PL inverse

$$
x = c - b\,\ln\!\left(\left(\frac{d - a}{y - a}\right)^{1/g} - 1\right)
\tag{inv-5PL}
$$

### Log-logistic 4PL inverse

$$
x = \frac{c}{\displaystyle\left(\frac{d - y}{y - a}\right)^{1/b}}
\tag{inv-LL4}
$$

### Log-logistic 5PL inverse

$$
x = c - \frac{1}{b}\,\ln\!\frac{\left(\dfrac{y-a}{d-a}\right)^{-g} - 1}{g}
\tag{inv-LL5}
$$

### Gompertz inverse

$$
x = c - \frac{1}{b}\,\ln\!\left(-\ln\frac{y - a}{d - a}\right)
\tag{inv-G4}
$$

## Worked example: round-trip verification

A useful sanity check is to confirm $x = f^{-1}(f(x))$ to machine
precision.

```{r inverse-roundtrip}
x_true <- 2.0  # log10(100 AU/mL)

# ---- 4PL ----
y_fwd  <- logistic4(x_true, p4$a, p4$b, p4$c, p4$d)
x_back <- inv_logistic4(y_fwd, p4$a, p4$b, p4$c, p4$d)
cat(sprintf("4PL   round-trip error: %.2e\n", abs(x_back - x_true)))

# ---- 5PL ----
y_fwd  <- logistic5(x_true, p5$a, p5$b, p5$c, p5$d, p5$g)
x_back <- inv_logistic5(y_fwd, p5$a, p5$b, p5$c, p5$d, p5$g)
cat(sprintf("5PL   round-trip error: %.2e\n", abs(x_back - x_true)))

# ---- loglogistic4 ----
x_ll4  <- log10(100)  # = 2.0; x_conc must be positive
y_fwd  <- loglogistic4(100, p_ll4$a, p_ll4$b, p_ll4$c, p_ll4$d)
x_back <- inv_loglogistic4(y_fwd, p_ll4$a, p_ll4$b, p_ll4$c, p_ll4$d)
cat(sprintf("LL4   round-trip error: %.2e\n", abs(x_back - 100)))

# ---- loglogistic5 ----
y_fwd  <- loglogistic5(x_true, p_ll5$a, p_ll5$b, p_ll5$c, p_ll5$d, p_ll5$g)
x_back <- inv_loglogistic5(y_fwd, p_ll5$a, p_ll5$b, p_ll5$c, p_ll5$d, p_ll5$g)
cat(sprintf("LL5   round-trip error: %.2e\n", abs(x_back - x_true)))

# ---- Gompertz ----
y_fwd  <- gompertz4(x_true, p_g4$a, p_g4$b, p_g4$c, p_g4$d)
x_back <- inv_gompertz4(y_fwd, p_g4$a, p_g4$b, p_g4$c, p_g4$d)
cat(sprintf("G4    round-trip error: %.2e\n", abs(x_back - x_true)))
```

All errors should be at or below $10^{-12}$, confirming numerical
consistency between the forward and inverse functions.

## Effect of fixing $a$ on back-calculated concentration

When the lower asymptote $a$ is fixed at a known value (e.g.\ the mean of
blank wells), the resulting inverse uses fewer estimated parameters and
yields narrower confidence intervals.

```{r fixed-a-demo}
# Suppose we know the blank mean is 95 MFI
a_fixed <- 95
y_test  <- 10000  # an observed sample MFI

x_free  <- inv_logistic4(y_test, p4$a, p4$b, p4$c, p4$d)
x_fixed <- inv_logistic4_fixed(y_test, a_fixed, p4$b, p4$c, p4$d)

cat(sprintf("Back-calculated x (a free):  %.4f\n", x_free))
cat(sprintf("Back-calculated x (a fixed): %.4f\n", x_fixed))
```

---

# Derivatives {#sec-derivatives}

## First derivatives (dy/dx)

All first derivatives are positive when $b > 0$ and $a < d$, confirming
strict monotonicity.

### 4PL — `dydx_logistic4`

Let $u = \exp(-(x-c)/b)$.  Then:

$$
\frac{dy}{dx} = \frac{(d-a)\,u}{b\,(1+u)^2}
\tag{D1-4PL}
$$

The maximum slope occurs at $x = c$ (the inflection point), where $u = 1$:

$$
\left.\frac{dy}{dx}\right|_{x=c} = \frac{d - a}{4b}
\tag{D1-4PL-inf}
$$

### 5PL — `dydx_logistic5`

$$
\frac{dy}{dx} = \frac{g\,(d-a)\,u}{b\,(1+u)^{g+1}}
\tag{D1-5PL}
$$

The 5PL inflection is at $x_{\inf} = c + b\ln g$, so the maximum slope is:

$$
\left.\frac{dy}{dx}\right|_{x=x_{\inf}}
= \frac{(d-a)\,g^{g/(g+1) - 1/g}}{b\,(1 + g^{-1})^{g+1}}
$$

### Log-logistic 4PL — `dydx_loglogistic4`

Let $r = (c/x)^b$.  Then:

$$
\frac{dy}{dx} = \frac{b\,(d-a)\,r}{x\,(1+r)^2}
\tag{D1-LL4}
$$

The inflection on the raw scale is at $x = c$, where $r = 1$:

$$
\left.\frac{dy}{dx}\right|_{x=c} = \frac{d-a}{4c}
$$

### Log-logistic 5PL — `dydx_loglogistic5`

Let $u = 1 + g\,e^{-b(x-c)}$.  Then:

$$
\frac{dy}{dx} = b\,(d-a)\,e^{-b(x-c)}\,u^{-1/g - 1}
\tag{D1-LL5}
$$

### Gompertz — `dydx_gompertz4`

Let $v = \exp(-b(x-c))$.  Then:

$$
\frac{dy}{dx} = b\,(d-a)\,v\,\exp(-v)
\tag{D1-G4}
$$

The Gompertz inflection is at $x = c$ ($v = 1$):

$$
\left.\frac{dy}{dx}\right|_{x=c} = \frac{b\,(d-a)}{e}
$$

## Second derivatives (d²y/dx²)

Second derivatives locate the inflection points and characterise the
curvature of each shoulder.
Analytical forms exist for the 4PL and Gompertz; numerical
central-difference approximations are used for the three remaining models.

### 4PL (analytical) — `d2x_logistic4`

$$
\frac{d^2y}{dx^2} = -\frac{(d-a)\,u\,(1-u)}{b^2\,(1+u)^3}
\tag{D2-4PL}
$$

At $x = c$ ($u = 1$) the second derivative is zero, confirming the
inflection.

### Gompertz (analytical) — `d2x_gompertz4`

$$
\frac{d^2y}{dx^2} = b^2\,(d-a)\,v\,(v-1)\,\exp(-v)
\tag{D2-G4}
$$

At $x = c$ ($v = 1$): $d^2y/dx^2 = 0$.
Consistent with $x = c$ being the Gompertz inflection point.

### 5PL, LL4, LL5 (numerical) — `d2x_logistic5`, `d2x_loglogistic4`, `d2x_loglogistic5`

For models where closed-form second derivatives are algebraically
unwieldy, a symmetric central-difference scheme is used with step size
$h = 10^{-5}$:

$$
\frac{d^2y}{dx^2}\bigg|_x \approx
\frac{f(x+h) - 2f(x) + f(x-h)}{h^2}
$$

This yields $O(h^2)$ accuracy; for typical parameter magnitudes the
approximation error is below $10^{-7}$.

## Worked example: slope at an arbitrary point and at the inflection

```{r derivatives-demo}
x_eval <- 2.0  # evaluate at log10(100) AU/mL

# ---- 4PL first derivative ----
slope_4pl <- dydx_logistic4(x_eval, p4$a, p4$b, p4$c, p4$d)
slope_inf  <- (p4$d - p4$a) / (4 * p4$b)   # analytical inflection slope
cat(sprintf("4PL  dy/dx at x=2.0  : %.2f MFI / log10-unit\n", slope_4pl))
cat(sprintf("4PL  dy/dx at x=c=%.1f: %.2f (formula: (d-a)/(4b))\n",
            p4$c, slope_inf))

# ---- Gompertz first derivative at inflection ----
slope_g4     <- dydx_gompertz4(p_g4$c, p_g4$a, p_g4$b, p_g4$c, p_g4$d)
slope_g4_inf <- p_g4$b * (p_g4$d - p_g4$a) / exp(1)
cat(sprintf("\nGompertz dy/dx at inflection (x=c): %.2f\n", slope_g4))
cat(sprintf("Gompertz formula b*(d-a)/e        : %.2f\n", slope_g4_inf))

# ---- Visualise slope profile across the curve ----
slopes_4pl  <- dydx_logistic4(x_seq, p4$a, p4$b, p4$c, p4$d)
slopes_g4   <- dydx_gompertz4(x_seq, p_g4$a, p_g4$b, p_g4$c, p_g4$d)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(x_seq, slopes_4pl, type = "l", lwd = 2, col = "#2166AC",
     xlab = expression(x == log[10](conc)),
     ylab = "dy/dx  (MFI / log10-unit)",
     main = "4PL slope profile")
abline(v = p4$c, lty = 2, col = "grey50")

plot(x_seq, slopes_g4, type = "l", lwd = 2, col = "#BF812D",
     xlab = expression(x == log[10](conc)),
     ylab = "dy/dx  (MFI / log10-unit)",
     main = "Gompertz slope profile")
abline(v = p_g4$c, lty = 2, col = "grey50")

par(mfrow = c(1, 1))
```

```{r d2-demo}
# Confirm d²y/dx² = 0 at the 4PL inflection
d2_at_inf <- d2x_logistic4(p4$c, p4$a, p4$b, p4$c, p4$d)
cat(sprintf("4PL d²y/dx² at inflection: %.2e  (expect 0)\n", d2_at_inf))

# Numerical second derivative for the 5PL
d2_5pl_at_c <- d2x_logistic5(p5$c, p5$a, p5$b, p5$c, p5$d, p5$g)
cat(sprintf("5PL d²y/dx² at x=c      : %.4f\n", d2_5pl_at_c))
# Note: the 5PL inflection is at x = c + b*log(g), not at x = c
x_inf_5pl <- p5$c + p5$b * log(p5$g)
d2_5pl_at_xinf <- d2x_logistic5(x_inf_5pl, p5$a, p5$b, p5$c, p5$d, p5$g)
cat(sprintf("5PL d²y/dx² at x_inf=%.3f: %.2e  (expect ~0)\n",
            x_inf_5pl, d2_5pl_at_xinf))
```

---

# Gradient Factory: `make_inv_and_grad_fixed` {#sec-gradients}

## Purpose

Delta-method confidence intervals for back-calculated concentrations
require three quantities for each observed response $y$:

1. $\hat{x} = f^{-1}(y;\,\hat\theta)$ — the point estimate.
2. $\nabla_\theta\, x = \partial x/\partial \theta$ — sensitivity of $x$
   to each model parameter.
3. $\partial x/\partial y$ — sensitivity of $x$ to measurement error
   in $y$.

The uncertainty in $\hat{x}$ is then:

$$
\operatorname{Var}(\hat{x})
\approx
\bigl(\nabla_\theta\, x\bigr)^\top
\,\mathbf{V}_{\hat\theta}\,
\bigl(\nabla_\theta\, x\bigr)
\;+\;
\left(\frac{\partial x}{\partial y}\right)^2\!\sigma_y^2
\tag{DeltaVar}
$$

where $\mathbf{V}_{\hat\theta} = \operatorname{Cov}(\hat\theta)$ is
the parameter covariance matrix and $\sigma_y^2$ is the assay noise
variance at $y$.

## API

```r
fns <- make_inv_and_grad_fixed(model, fixed_a = NULL)
```

**Arguments**

- `model` — character string, one of `"logistic4"`, `"logistic5"`,
  `"loglogistic4"`, `"loglogistic5"`, `"gompertz4"`.
- `fixed_a` — numeric scalar or `NULL`.  If non-`NULL`, $a$ is treated
  as a known constant and excluded from $\nabla_\theta\, x$,
  typically reducing the number of free parameters from 4 (or 5) to
  3 (or 4).

**Return value**

A list of three closures, each with signature `function(y, p)` where
`p` is the named vector returned by `coef(fit)`:

| Closure | Returns |
|---------|---------|
| `$inv(y, p)` | Scalar $\hat x$ |
| `$grad(y, p)` | Named numeric vector $\nabla_\theta\, x$ |
| `$grad_y(y, p)` | Scalar $\partial x/\partial y$ |

## Analytical gradient formulae

### 4PL

Using $x = c + b\ln\!\left(\frac{y-a}{d-y}\right)$:

$$
\frac{\partial x}{\partial a} = \frac{-b}{y-a}, \quad
\frac{\partial x}{\partial b} = \ln\frac{y-a}{d-y}, \quad
\frac{\partial x}{\partial c} = 1, \quad
\frac{\partial x}{\partial d} = \frac{-b}{d-y}
$$

$$
\frac{\partial x}{\partial y} = \frac{b\,(d-a)}{(y-a)(d-y)}
$$

### 5PL

Let $T = (d-a)/(y-a)$, $T_g = T^{1/g}$, $W = T_g - 1$:

$$
\frac{\partial x}{\partial a} = \frac{b\,T_g}{g\,W\,(y-a)}, \quad
\frac{\partial x}{\partial b} = -\ln W, \quad
\frac{\partial x}{\partial c} = 1, \quad
\frac{\partial x}{\partial d} = \frac{-b\,T_g}{g\,W\,(d-a)}
$$

$$
\frac{\partial x}{\partial g} = \frac{b\,T_g\,\ln T}{g^2\,W}, \quad
\frac{\partial x}{\partial y} = \frac{-b\,T_g}{g\,W\,(y-a)}
$$

### Log-logistic 4PL

Let $Q = (d-y)/(y-a)$, $p = 1/b$, $x = c/Q^p$:

$$
\frac{\partial x}{\partial a} = \frac{-c p\,(d-y)\,Q^{-p-1}}{(y-a)^2}, \quad
\frac{\partial x}{\partial b} = \frac{x\,\ln Q}{b^2}, \quad
\frac{\partial x}{\partial c} = Q^{-p}, \quad
\frac{\partial x}{\partial d} = c p\,Q^{-p-1}/(y-a)
$$

$$
\frac{\partial x}{\partial y} = \frac{c p\,(d-a)\,Q^{-p-1}}{(y-a)^2}
$$

### Gompertz

Let $R = (y-a)/(d-a)$, $L = -\ln R > 0$:

$$
\frac{\partial x}{\partial a} = \frac{1}{b\,L\,(y-a)}, \quad
\frac{\partial x}{\partial b} = \frac{\ln L}{b^2}, \quad
\frac{\partial x}{\partial c} = 1, \quad
\frac{\partial x}{\partial d} = \frac{-1}{b\,L\,(d-a)}
$$

$$
\frac{\partial x}{\partial y} = \frac{1}{b\,L\,(y-a)}
$$

### Log-logistic 5PL

Let $\rho = (y-a)/(d-a)$, $\rho_g = \rho^{-g}$, $V = \rho_g - 1$:

$$
\frac{\partial x}{\partial a} = \frac{\rho^{-g-1}}{b\,g\,V\,(d-a)}, \quad
\frac{\partial x}{\partial b} = \frac{\ln V - \ln g}{b^2}, \quad
\frac{\partial x}{\partial c} = 1
$$

$$
\frac{\partial x}{\partial d} = \frac{\rho^{-g-1}(y-a)}{b\,g\,V\,(d-a)^2}, \quad
\frac{\partial x}{\partial g} = \frac{1 - \ln(g V)}{b\,g^2}
$$

$$
\frac{\partial x}{\partial y} = \frac{\rho^{-g-1}}{b\,g\,V\,(d-a)}
$$

## Worked example: gradient factory in action

```{r gradient-factory}
# Simulate a coefficient vector as would come from coef(fit)
p_hat <- c(a = 100, b = 0.8, c = 1.5, d = 50000)

# Build the closure set for 4PL with a estimated freely
fns_free <- make_inv_and_grad_fixed("logistic4", fixed_a = NULL)

y_obs <- 15000  # observed MFI

x_hat  <- fns_free$inv(y_obs, p_hat)
g_hat  <- fns_free$grad(y_obs, p_hat)
gy_hat <- fns_free$grad_y(y_obs, p_hat)

cat(sprintf("Back-calculated log10(conc): %.4f\n", x_hat))
cat("Gradient w.r.t. theta:\n")
print(round(g_hat, 5))
cat(sprintf("Gradient w.r.t. y: %.6f\n", gy_hat))

# ---- Verify against finite differences ----
eps <- 1e-6
fd_grad <- sapply(names(p_hat), function(nm) {
  p_up       <- p_hat
  p_up[[nm]] <- p_hat[[nm]] + eps
  (fns_free$inv(y_obs, p_up) - x_hat) / eps
})
cat("\nFinite-difference check (should match grad above):\n")
print(round(fd_grad, 5))

# ---- With fixed_a ----
fns_fixed <- make_inv_and_grad_fixed("logistic4", fixed_a = 95)
g_fixed   <- fns_fixed$grad(y_obs, p_hat[c("b", "c", "d")])
cat("\nGradient with fixed a=95 (b, c, d only):\n")
print(round(g_fixed, 5))
```

### Applying the delta-method variance formula

```{r delta-method}
# A toy parameter covariance matrix (realistic order-of-magnitude)
V_theta <- diag(c(a = 4, b = 0.002, c = 0.01, d = 250000))
rownames(V_theta) <- colnames(V_theta) <- names(p_hat)

sigma_y <- 200   # assay CV ≈ 1.3% at MFI=15000

var_x_param <- as.numeric(t(g_hat) %*% V_theta %*% g_hat)
var_x_meas  <- gy_hat^2 * sigma_y^2
var_x_total <- var_x_param + var_x_meas

cat(sprintf("Variance from parameter uncertainty: %.5f (log10 units^2)\n",
            var_x_param))
cat(sprintf("Variance from measurement noise    : %.5f\n", var_x_meas))
cat(sprintf("Total variance in back-calculated x: %.5f\n", var_x_total))
cat(sprintf("95%% CI half-width: ± %.4f log10 units\n",
            qnorm(0.975) * sqrt(var_x_total)))
```

---

# Prediction Grid Utilities {#sec-grid}

## `generate_prediction_grid`

The prediction grid is a data frame of concentration values at which
the fitted curve and its confidence band are evaluated for plotting and
interpolation.
Both **curveRfreq** and **curveRbayes** use the same grid structure,
ensuring visual consistency.

```r
generate_prediction_grid(
  std_curve_conc,
  n_grid             = 200L,
  grid_min_conc      = 1e-4,
  grid_max_conc      = NULL,
  is_log_independent = TRUE
)
```

**Key arguments**

- `std_curve_conc` — the maximum standard concentration (AU/mL).
  When `grid_max_conc = NULL`, this value is used as the upper boundary.
- `n_grid` — number of evenly-spaced grid points.  Default 200 gives
  smooth curves without excessive computation.
- `grid_min_conc` — lower boundary of the grid on the raw scale.
  Must be $> 0$.
- `is_log_independent` — if `TRUE` (default), grid points are
  evenly spaced on the $\log_{10}$ scale, matching the fitting scale
  for log-transformed models; if `FALSE`, points are evenly spaced on
  the linear concentration scale.

**Return value**

A data frame with three columns:

| Column | Description |
|--------|-------------|
| `log10_concentration` | $\log_{10}$ of concentration |
| `concentration` | Concentration on the raw (AU/mL) scale |
| `x_fit` | The value actually passed to the model ($= \log_{10}c$ when `is_log_independent = TRUE`) |

## `predict_grid_response`

Evaluates the forward model at every grid point:

```r
predict_grid_response(grid, model_name, params)
```

Returns a numeric vector of $\hat y$ values of length `n_grid`.

## Worked example: building and plotting a prediction grid

```{r prediction-grid}
# Generate a 300-point log-scale grid from 0.0001 to 30 AU/mL
grid <- generate_prediction_grid(
  std_curve_conc = 30,
  n_grid         = 300,
  grid_min_conc  = 1e-4,
  is_log_independent = TRUE
)

cat("Grid dimensions:", nrow(grid), "×", ncol(grid), "\n")
cat("x_fit range    :", round(range(grid$x_fit), 3), "\n")
cat("conc range     :", signif(range(grid$concentration), 3), "AU/mL\n")

# Evaluate all five models on the grid
params_4pl  <- c(a = 100, b = 0.8, c = 1.5, d = 50000)
params_5pl  <- c(a = 100, b = 0.8, c = 1.5, d = 50000, g = 0.6)
params_ll4  <- c(a = 100, b = 1.8, c = 30,  d = 50000)
params_ll5  <- c(a = 100, b = 1.2, c = 1.5, d = 50000, g = 0.5)
params_g4   <- c(a = 100, b = 1.2, c = 1.5, d = 50000)

yhat_4pl  <- predict_grid_response(grid, "logistic4",    params_4pl)
yhat_5pl  <- predict_grid_response(grid, "logistic5",    params_5pl)
yhat_ll5  <- predict_grid_response(grid, "loglogistic5", params_ll5)
yhat_g4   <- predict_grid_response(grid, "gompertz4",    params_g4)

# For loglogistic4 the grid x_fit must be on the raw (positive) scale
grid_lin <- generate_prediction_grid(
  std_curve_conc = 500, n_grid = 300,
  grid_min_conc = 0.01, is_log_independent = FALSE
)
yhat_ll4 <- predict_grid_response(grid_lin, "loglogistic4", params_ll4)
```

```{r grid-plot, fig.height = 5}
cols <- c("4PL"  = "#2166AC", "5PL"  = "#D6604D",
          "LL5"  = "#762A83", "G4"   = "#BF812D")

plot(grid$x_fit, yhat_4pl, type = "l", lwd = 2, col = cols["4PL"],
     xlab = expression(log[10](concentration~~"(AU/mL)")),
     ylab = "Response (MFI)",
     main = "Prediction grids — all five models",
     ylim = c(0, 55000))
lines(grid$x_fit, yhat_5pl,  lwd = 2, col = cols["5PL"])
lines(grid$x_fit, yhat_ll5,  lwd = 2, col = cols["LL5"])
lines(grid$x_fit, yhat_g4,   lwd = 2, col = cols["G4"])
lines(log10(grid_lin$concentration), yhat_ll4,
      lwd = 2, col = "#4DAC26", lty = 2)

legend("topleft", bty = "n", lwd = 2,
       legend = c("logistic4", "logistic5 (g=0.6)", "loglogistic5 (g=0.5)",
                  "gompertz4", "loglogistic4"),
       col    = c(cols, "#4DAC26"),
       lty    = c(1, 1, 1, 1, 2))
```

## Using the grid for back-calculation across a response range

```{r grid-backcalc}
# Demonstrate back-calculation at a sequence of response values
y_targets <- seq(500, 45000, by = 2000)

x_back <- inv_logistic4(y_targets, params_4pl["a"], params_4pl["b"],
                        params_4pl["c"], params_4pl["d"])

result <- data.frame(
  MFI            = y_targets,
  log10_conc     = round(x_back, 4),
  conc_AU_mL     = round(10^x_back, 3)
)
print(result)
```

---

# Model Selection Considerations {#sec-selection}

The table below summarises the key mathematical properties that should
guide model choice for a given assay.

| Property | 4PL | 5PL | LL4 | LL5 | Gompertz |
|----------|-----|-----|-----|-----|----------|
| Inflection symmetry | Symmetric | Asymmetric | Symmetric | Asymmetric | Asymmetric (fixed) |
| Inflection at $x = c$? | Yes | Only if $g=1$ | Yes | No | Yes |
| Extra shape parameter | No | Yes ($g$) | No | Yes ($g$) | No |
| Closed-form $d^2y/dx^2$ | Yes | No | No | No | Yes |
| Raw-scale $x$ required | No | No | Yes | No | No |
| Typical application | Log-linear ELISA | Asymmetric ELISA | Hill-equation dose–response | Flexible dose–response | Left-skewed saturation |

**Practical guidance:**

- Start with `logistic4`.  It has the fewest parameters and closed-form
  derivatives.
- If the residuals show systematic skew in one shoulder, add the
  asymmetry parameter ($g$) via `logistic5`.
- Use `loglogistic4` / `loglogistic5` when you prefer to work on the
  raw concentration scale (e.g.\ for pharmacological IC50 reporting).
- `gompertz4` is a natural choice for bead-based multiplex assays where
  the response rises steeply from a low baseline and saturates slowly.

---

# Numerical Precision Notes {#sec-precision}

1. **Step size for numerical second derivatives.**  The default $h = 10^{-5}$
   gives $O(h^2) \approx 10^{-10}$ truncation error for parameters of
   order unity.  Reduce $h$ if parameter magnitudes exceed $10^4$.

2. **Asymptote guard in inverse functions.**  `inv_logistic4` and
   `inv_logistic5` apply a tolerance buffer of $10^{-6}$ and return `NA`
   for responses within that distance of either asymptote, where the
   logit transformation diverges.

3. **Overflow protection in Gompertz.**  The data-generation helper in
   `bead_assay_example.R` caps the exponent $u = b(\ln x - \ln c)$ at
   30 to prevent `exp(exp(u))` overflow for extreme concentrations; the
   core `gompertz4` function does not apply this cap.

4. **Machine precision of round-trips.**  As demonstrated in
   Section \@ref(sec-inverses), forward–inverse round-trips return errors
   at the level of $\sim 10^{-13}$, consistent with IEEE 754
   double-precision arithmetic.

---

# References

<div id="refs"></div>

---

# Session Information

```{r session-info}
sessionInfo()
```

---

